---
sstitle: "Final Project"
author: "William Golden"
date: "3/29/2022"
output: html_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r,echo=FALSE,message=FALSE,results='hide',warning=FALSE}
library(MASS)
library(forecast)
library(tidyverse)
library(prophet)
library(reshape2)
library(leaflet)
library(gganimate)


df1 <- read.csv("./Data/Raw/2000-2009.csv")

df1$YEAR <- df1$DATE %>% str_trunc(4,ellipsis = "") %>% as.numeric()

Clean1 <- df1 %>% 
  group_by(YEAR,NAME, LATITUDE,LONGITUDE) %>% 
  summarize(TotalPrecip=sum(PRCP,na.rm = TRUE),
            AvgPrecip=mean(PRCP,na.rm=TRUE))

df2 <- read.csv("./Data/Raw/2010-Present.csv")

df2$YEAR <- df2$DATE %>% str_trunc(4,ellipsis = "") %>% as.numeric()

Clean2 <- df2 %>% 
  group_by(YEAR,NAME, LATITUDE,LONGITUDE) %>% 
  summarize(TotalPrecip=sum(PRCP,na.rm = TRUE),
            AvgPrecip=mean(PRCP,na.rm=TRUE))

df3 <- read.csv("./Data/Raw/1990-1999.csv")

df3$YEAR <- df3$DATE %>% str_trunc(4,ellipsis = "") %>% as.numeric()

Clean3 <- df3 %>% 
  group_by(YEAR,NAME, LATITUDE,LONGITUDE) %>% 
  summarize(TotalPrecip=sum(PRCP,na.rm = TRUE),
            AvgPrecip=mean(PRCP,na.rm=TRUE))

df4 <- read.csv("./Data/Raw/1980-1989.csv")

df4$YEAR <- df4$DATE %>% str_trunc(4,ellipsis = "") %>% as.numeric()

Clean4 <- df4 %>% 
  group_by(YEAR,NAME, LATITUDE,LONGITUDE) %>% 
  summarize(TotalPrecip=sum(PRCP,na.rm = TRUE),
            AvgPrecip=mean(PRCP,na.rm=TRUE))



df5 <- read.csv("./Data/Raw/1970-1979.csv")

df5$YEAR <- df5$DATE %>% str_trunc(4,ellipsis = "") %>% as.numeric()

Clean5 <- df5 %>% 
  group_by(YEAR,NAME, LATITUDE,LONGITUDE) %>% 
  summarize(TotalPrecip=sum(PRCP,na.rm = TRUE),
            AvgPrecip=mean(PRCP,na.rm=TRUE))



Clean <- rbind(Clean1,Clean2,Clean3,Clean4,Clean5)

rm(list=setdiff(ls(), "Clean"))
```


# Introduction


```{r,echo=FALSE,message=FALSE,results='hide',warning=FALSE}
library(sf)
library(leaflet)
devtools::install_github("ropensci/USAboundaries")
devtools::install_github("ropensci/USAboundariesData")
library(USAboundaries)

m <- leaflet() %>% 
  setView(lng = -111.6952, lat = 40.1542,zoom = 8) %>% 
  addTiles()

# extract polygon info for Utah County border
ut_counties <- us_counties(states = "Utah",resolution = "high") 

#clean extra column
ut_counties <- ut_counties[,-9]
ut_counties %>% 
  filter(name == "Utah")


utah_co <- ut_counties %>% 
  filter(name == "Utah")

ut_co_poly <- utah_co$geometry
latlon <- ut_co_poly %>% unlist()
long <- latlon[1:(length(latlon)/2)]
lat <- latlon[(length(long)+1):length(latlon)]

```


For this project, I wanted to look at weather data in Utah County (as seen above) and predict when it would rain based
off of the collected data. I gathered data from the 
National Center for Environmental Information (https://www.ncdc.noaa.gov/cdo-web/). This data shows the amount of precipitation recorded each day from 1970-2022. 


# Analyzing the Data

This data set contains a ton of daily records of precipitation in Utah County. After wrangling the data, 
the dataset looks like this: 
```{r}
Clean %>% glimpse()
```


To better analyze 
where this data is being collected, here are two maps showing the location of the weather stations 
and how much precipitation each one recorded for the year 1970 and 2021. 

```{r, message=FALSE,echo=FALSE}
stations <- Clean %>% filter(YEAR==1970)

stations_long <- stations$LONGITUDE
stations_lat <- stations$LATITUDE


m %>% 
  addPolygons(lng = long,
              lat = lat,
              color = "Grey",
              fillOpacity = .1) %>% 
  addCircles(lng = stations_long,
             lat=stations_lat,
             radius = stations$AvgPrecip*40000,
             label = paste0(stations$NAME,": Total Precipiation (inches) in 1970= ",
                            stations$TotalPrecip),
             fillColor = "Blue",
             color = "Purple")

stations2 <- Clean %>% filter(YEAR==2021)


stations_long2 <- stations2$LONGITUDE
stations_lat2 <- stations2$LATITUDE


m %>% 
  addPolygons(lng = long,
              lat = lat,
              color = "Grey",
              fillOpacity = .1) %>% 
  addCircles(lng = stations_long2,
             lat=stations_lat2,
             radius = stations2$AvgPrecip*40000,
             label = paste0(stations2$NAME,": Total Precipiation (inches) in 2021= ",
                            stations2$TotalPrecip),
             fillColor = "Blue",
             color = "Purple")
```


Something I wanted to look for in this dataset is how precipitation has changed over time in Utah County.
I created this graph showing the change in total precipitation in Utah County from 1970 until now. 

```{r,message=FALSE,echo=FALSE}
q <- Clean %>% 
  group_by(YEAR) %>% 
  summarize(TotalPrcp = mean(TotalPrecip)) %>% 
  ggplot(aes(x=YEAR,y=TotalPrcp))+
  geom_line() +
  labs(x="Year",
       y="Total Precipiation (Inches)",
       title = "Total Precipiation over Time in Utah County")
q + 
  geom_point() +
  transition_reveal(YEAR)+
  theme_classic()
```


Another visual I've created is the variance of yearly recorded precipitation across different
weather stations. The following graph shows how much change is in the total precipitation recorded
for the weather stations each year. 

```{r, echo=FALSE,message=FALSE}
p <- Clean %>% 
  group_by(YEAR, NAME) %>% 
  summarize(TotalPrcp = mean(TotalPrecip)) %>% 
  ggplot(aes(x=NAME, y=TotalPrcp))+
  geom_col()+
  labs(x="Station Name",
       y="Total Precipiation (Inches)")+
  theme(axis.text.x = (element_text(angle = 60, hjust = 1)))


p+transition_time(YEAR) +
  labs(title = "Year: {frame_time}")
```


## Creating a Predictive Model Using the Prohpet Package

I fed my data into a program called Prophet (made by Facebook) to create predictions of future
amounts of precipitation based off historical data. 

```{r,message=FALSE,echo=FALSE,warning=FALSE}
df1 <- read.csv("./Data/Raw/2000-2009.csv")

Clean1 <- df1 %>% 
  group_by(DATE,NAME) %>% 
  summarize(TotalPrecip=sum(PRCP,na.rm = TRUE),
            AvgPrecip=mean(PRCP,na.rm=TRUE))

df2 <- read.csv("./Data/Raw/2010-Present.csv")

Clean2 <- df2 %>% 
  group_by(DATE,NAME) %>% 
  summarize(TotalPrecip=sum(PRCP,na.rm = TRUE),
            AvgPrecip=mean(PRCP,na.rm=TRUE))

df3 <- read.csv("./Data/Raw/1990-1999.csv")

Clean3 <- df3 %>% 
  group_by(DATE,NAME) %>% 
  summarize(TotalPrecip=sum(PRCP,na.rm = TRUE),
            AvgPrecip=mean(PRCP,na.rm=TRUE))

df4 <- read.csv("./Data/Raw/1980-1989.csv")

Clean4 <- df4 %>% 
  group_by(DATE,NAME) %>% 
  summarize(TotalPrecip=sum(PRCP,na.rm = TRUE),
            AvgPrecip=mean(PRCP,na.rm=TRUE))



df5 <- read.csv("./Data/Raw/1970-1979.csv")

Clean5 <- df5 %>% 
  group_by(DATE,NAME) %>% 
  summarize(TotalPrecip=sum(PRCP,na.rm = TRUE),
            AvgPrecip=mean(PRCP,na.rm=TRUE))



Clean <- rbind(Clean1,Clean2,Clean3,Clean4,Clean5)

df1 <- Clean %>% 
  select(TotalPrecip,DATE)

df1 <- rename(df1, ds=DATE)

df1 <- rename(df1, y=TotalPrecip)
```

```{r,message=FALSE,echo=FALSE,warning=FALSE}
m <- prophet(df1)

future <- make_future_dataframe(m, periods = 365)

forecast <- predict(m, future)

plot(m, forecast)
```

This doesn't look like much, but this graph is showing how much data the Prophet program is analyzing as 
well as a short prediction of future amounts of precipitation in Utah County. Since day-to-day, the 
amount of precipitation is between 0-1 inches, the data is hard to visualize in a large graph like this.


Another tool from the Prophet package is to show model-created trends in precipitation over the year and 
over time. 

```{r,message=FALSE,echo=FALSE,warning=FALSE}
prophet_plot_components(m, forecast)
```




## Conclusion

I've managed to wrangle real-life data into a workable data set that shows the changes of precipitation
 in Utah County for over 50 years. I've used the Prophet package to predict the rainfall amounts in the 
 following year as well as displayed how the Prophet packages learns the rainfall patterns in Utah County. 


